---
title: "Introduction to Data Visualization"
author: "Ketura Elie"
date: "2023-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, load_packages}
library(tidyverse)
```

## Anscombe 

```{r,a_1}
anscombe %>% glimpse() #not tidy data table, columns contain 2 piece of information
```

## Tidy 

Tidy the `anscombe` data set.  

```{r, tt_1}
a_tidy <- anscombe %>% 
    tibble::rowid_to_column() %>%
    pivot_longer(cols = !c("rowid")) %>% 
    tidyr::separate(name,
                    c("variable","dataset"),
                    sep = 1) %>% 
    pivot_wider(names_from='variable',values_from = 'value') %>% 
    arrange(dataset)
```

The tidy data long format data 

```{r, tt_2}
a_tidy %>% glimpse()
```


## Visualization

The 3 major components of any (statistical) figure.  
 
 * data - a table (dataframe)  
 * aesthetics - what do we want to see (the columns from a table)  
 * geometric objects - "how" do we want to see the variables  
 


## Aesthetics

```{r}
a_tidy %>% 
    ggplot(mapping = (aes(x=x,y=y))) +
    geom_point(mapping = aes(alpha = x))
```


Use facets to show the different groups

```{r, tt_3}
a_tidy %>% 
    ggplot(aes(x=x,y=y)) +
    geom_point() +
    facet_wrap(~dataset, labeller = 'label_both')
```



 We can easily map any aesthetic to the `dataset` column.  
 
```{r, tt_4}
a_tidy %>% 
    ggplot(aes(x=x,y=y)) +
    geom_point(aes(color = dataset,
                   shape = dataset),
               size = 3.5) +
    facet_wrap(~dataset, labeller = 'label_both', scales = "free") +
    theme(legend.position='none')
```
 
  
 We can easily add in best fit line with another layer.  
 
```{r, tt_5}
a_tidy %>% 
    ggplot(aes(x=x,y=y)) +
    geom_point(aes(color =dataset,
                   shape = dataset),
               size = 3.5) +
    geom_smooth(method=lm,
                formula = y ~ x,
                mapping = aes(color = dataset,
                               fill = dataset)) +
    facet_wrap(~dataset, labeller = 'label_both') +
    labs(title = "Examining Anscombe") +
    theme_bw() +
    theme(legend.position='none',
          plot.title = element_text(hjust = 0.5)) 
```
 